---
title: "A Field Guide to Lies & Stats - Levitin 2016*"
date: "2019-08-10"
output:
  blogdown::html_page:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(tidyverse)

# Data
product_ratings <- tibble(Product = c("Mine", "Competitor's"),
                          mean_rating = c(71, 68))
```

If you want to see the full code for this post then you can find the .Rmd file [here](https://github.com/mitchelfruin/MitchThinks/blob/master/content/summaries/lies_and_stats_Levitin_2016.Rmd).

# Summary

Levitin provides a solid compilation of mistakes to avoid and tricks to look out for when producing and reading research respectively. However, given his background in  psychology and neuroscience, I was hoping for a more general thread pulling the examples together concerning *why* the human brain is 'fooled' by particular ways of presenting information rather than just *what* those presentations are.

The book's main strength was illustrative examples some of which stuck in my mind and comprise the rest of this post.

# Is the stat plausible?

The first thing to check when presented with a statistic, especially out of context, is whether the claim is even vaguely reasonable. 

## Back of the napkin calculations

One way to check plausibility is a quick on-the-fly calculation.

Say, for example, we run into someone claiming that "in the 35 years since marijuana laws stopped being enforced in California, the number of marijuana smokers has doubled every year". 

Well, even if we start with the most generous estimate that initially there was only 1 lonely pothead in California, then if that number doubled every year for 35 years:

$$\begin{aligned}
\text{Year 1:} \;\; 1 &= 2^0 \\
\text{Year 2:} \;\; 2 &= 2^1 \\
\text{Year 3:} \;\; 4 &= 2^2 \\
\vdots \\
\text{Year 35:} \;\; &= 2^{34} \\
\end{aligned}$$

So let's see just how many marijuana smokers we've currently got if that stat holds:
```{r potheads}
# Too high to count that high
2^34
```

Just over 17.1 billion people, which seems more than a little high. 

## You've taken a bath before

In a maths lecture at uni the professor began by writing a complicated looking formula on the board and told us that it described the time taken for a bath to empty as a function of its volume and the diameter of the plughole. He then asked us to raise our hand if we believed the bath would ever empty. When only a few people sheepishly did so he shouted up into the lecture theatre, "have you fucking taken a bath before?". His point was, just because we're dealing with supposedly scary maths doesn't mean you should ignore common sense. You've taken a bath before, you know the answer, whether you understand the formula or not. 

These kind of 'wait, but we're still talking about the real world' checks can sometimes help assess the plausibility of a statistic. For example, in 1993 New Jersey adopted a 'family cap' law that denied additional benefits to mothers who had children whilst already on welfare. After two months, legislators were celebrating its success as births had already fallen by 16%. The only problem being that human pregnancies generally take a bit longer than that. 

# What does the average mean?

## Mean, median and mode

## All averages aren't made equal

Relative percentage change vs. absolute percentage change (i.e. percentage points) vs. natural frequencies. 

## Bimodal distributions

Humans on average have one testicle. Meaningless to compute an average on a metric over a whole population where there are important differences. 

# Are the axes leading you astray?

Manipulating the axes on a graph is one of the easiest ways to con a reader. 

## Truncated

Shortening the proper range of y-axis values is often used to exaggerate the difference between two groups that are being compared.

For example, if I wanted to convince you that my product was way better than a competitor's I might show you a bar chart reporting the average customer rating like this:

```{r truncated bar chart, fig.align='center'}
# Bar chart with truncated y-axis
ggplot(product_ratings, aes(x = Product, y = mean_rating, fill = Product)) +
  geom_col() +
  scale_fill_discrete(guide = "none") +
  scale_y_continuous("Customer\nRating") +
  coord_cartesian(ylim = c(67.5, 71.5)) +
  theme_minimal() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```

It looks like my product blows my competitor's out of the water. However, notice where the y-axis starts and ends. 

If we're being strict, the y-axis should include the full range of values that customers were able to provide (it should at the very least include 0 if that's a possible value). Once we properly format the graph my product looks a little less impressive:

```{r full range bar chart, fig.align='center'}
# Bar chart with truncated y-axis
ggplot(product_ratings, aes(x = Product, y = mean_rating, fill = Product)) +
  geom_col() +
  scale_fill_discrete(guide = "none") +
  scale_y_continuous("Customer\nRating", limits = c(0, 100)) +
  theme_minimal() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```

## Discontinuous

Condensing a small portion of the x-axis can be used to falsely emphasise a few data points. 

For example, say I'm selling security systems and I want to put a scary graph on the leaflet I drop through people's letterboxes. I might include a line graph like this one that exaggerates the recent increase in the number of burglaries: 

```{r discontinuous line graph, fig.align='center'}
tibble(Year = c("2009", "2010", "2011", "2012", "2013", "2014", "2019"),
       burglary_count = c(100, 102.5, 105, 107.5, 110, 112.5, 125)) %>%
  ggplot(aes(x = Year, y = burglary_count)) +
  geom_line(group = 1) +
  scale_y_continuous("Burglary\nCount") +
  theme_minimal() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```

However, the apparent spike in burglaries is just an artefact of the fact that 5 year's worth of increases have been squeezed into one notch on the x-axis. 

## Elongated

Discontinuous axes make a change look more extreme by removing some values, but changes can also be exaggerated by adding in irrelevant blank values. 

For example, take a similar line graph to the one discussed above. If we plot the graph without any trickery then the data seem to show a steady increase in the number of burglaries: 

```{r normal line graph, fig.align='center'}
tibble(Year = c(2009:2019), burglary_count = c(seq(100, 112.5, 1.25))) %>%
  ggplot(aes(x = Year, y = burglary_count)) +
  geom_line(group = 1) +
  scale_y_continuous("Burglary\nCount") +
  scale_x_continuous(breaks = c(2009, 2011, 2013, 2015, 2017, 2019)) +
  theme_minimal() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```

However, we can make the same data look much more extreme by including space on the x-axis for years without any data. This narrows the visual space over which the increase in our data must be plotted meaning the line is steeper. Thus, the same increase over the same timespan looks much more drastic:

```{r elongated line graph, fig.align='center'}
tibble(Year = c(2009:2019), burglary_count = c(seq(100, 112.5, 1.25))) %>%
  ggplot(aes(x = Year, y = burglary_count)) +
  geom_line(group = 1) +
  scale_y_continuous("Burglary\nCount") +
  coord_cartesian(xlim = c(2004, 2024)) +
  theme_minimal() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```
Include axis space beyond the range of your data to make changes look more extreme. 

## Duplicate

If they measure the same thing then you can fudge the graph however you want. Only use if they're different units and you want to make within axis comparisons not between. E.g. planned parenthood Jason Chaffetz. 

# Dodgy reporting 

## Hiding a decrease in the cumulative

Tim Cook iPhone sales. 

## Unrelated plotting

Spurious time series. 

## Irrelevant differences

Just increase N enough to lower the p-value until it's 'significant'.

## Comparing apples and oranges 

**Amalgamating**: group samples that differ on an important dimension to make the stat more shocking. "50% of 10-18 year olds are sexually active". 

**Subdividing**: split things inconsistently to make your point more impressive. Divide all causes of death into subcategories apart from the one you want to promote. Should be equal bins (either by percentile of absolute).

