---
title: "Probability in the real world*"
date: "2019-08-15"
output:
  blogdown::html_page:
    toc: true
    toc_depth: 1
    number_sections: true
---

```{r setup, include=FALSE}
# Rmarkdown
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(tidyverse)
library(waffle) # Need to add Font Awesome using extrafont::
library(reshape2)
```

# Probability 101

## Probability vs. Statistics

Probability: Use known state of the world to predict the data we should observe.

* I have a fair coin (known state of the world) and can expect 50 heads and 50 tails, on average, from 100 flips (predicted data). 

Statistics: Use data we have observed to infer the most likely state of the world. 

* I flipped a coin 100 times and got 52 heads and 48 tails (observed data), I'm pretty sure that I have a fair coin (inferred state of the world). 

## Types of probability

Classic, based on symmetry. 

Frequentist, based on experience. 

Subjective, based on best guess. 

## Types of events

**Independent probabilities.** Should multiply. 

**Conditional probabilities.** Should use Bayes law. Visualise Bayesian updating with a contingency table/ infographic. 

## Puzzles

Sometimes probability is unintuitive. Birthday puzzle. 

# Legal problems

## Ignoring dependence

Sally Clark had two children dies from SIDS. Dr. Meadow, pediatrician, said there's 1 in 8543 chance of having a child die from SIDS giving a 1 in 73mil chance of having two children die from it. Potential genetic, environmental dependence ignored and she was sentenced for 3 years.

## The prosecutor's fallacy

**Prosecutor's fallacy:** False belief that conditional probabilities are invertible.

$$Pr(A|B) \neq Pr(B|A)$$

Forensics may compute that probability of blood at scene matching the defendent's by chance is 1%. That doesn't mean there's a 1% chance the defendent is innocent. We have been told the probability of a match given innocence, we want the probability of innocence given a match, but they don't equate:

$$Pr(\text{match}|\text{innocent}) \neq Pr(\text{innocent}|\text{match})$$

# Health problems

## Base rate neglect

Disease with 5% prevalence. Test is 95% accurate.

Let's unpack the two parts of the scenario using an example population of 400 people.

Part 1: "The disease has a prevalence of 5%":

* 5% of the 400 people have the disease $\rightarrow$ 20 people
* 95% of the 400 people don't $\rightarrow$ 380 people

Part 2: "The test is 95% accurate":

* 95% of the 20 people with the disease will be correctly identified as positive $\rightarrow$ 19 people
    + True positives
* 5% of the 20 people with the disease will be incorrectly identified as negative $\rightarrow$ 1 person
    + False negatives
* 95% of the 380 people without the disease will be correctly identified as negative $\rightarrow$ 361 people
    + True negatives
* 5% of the 380 people without the disease will be incorrectly identified as positive $\rightarrow$ 19 people
    + False positives

So, given a positive result the probability you have the disease is:

$$\begin{aligned}
\text{Pr}(D|+) &= \frac{\text{True pos.}}{\text{Total pos.}} \\
&= \frac{19}{19 + 19} \\
&= 0.5
\end{aligned}$$

```{r base rate neglect}
# Waffle chart 400 people
example <- c("True pos" = 19, "False pos" = 19, "True neg" = 361, "False neg" = 1)

waffle(example,
       rows = 20)

# Bar chart 1000 people
tibble(disease = c("Positive", "Negative"),
       test_pos = c(47.5, 47.5),
       test_neg = c(2.5, 902.5)) %>%
  melt(id.vars = "disease") %>%
  ggplot(aes(x = disease, y = value, fill = variable)) +
  geom_col() +
  scale_x_discrete("Disease\nStatus") +
  scale_y_continuous("No. of People", limits = c(0, 1000)) +
  coord_flip() +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.grid.major = element_line())
```

# Research problems

# Statistical cognition

## The linda problem

Extra example of how we aren't intuitive statisticians. 

## All averages aren't made equal

Relative percentage change vs. absolute percentage change (i.e. percentage points) vs. natural frequencies. 