---
title: "Probability in the real world*"
date: "2019-08-15"
output:
  blogdown::html_page:
    toc: true
    toc_depth: 1
    number_sections: true
---

```{r setup, include=FALSE}
# Rmarkdown
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# Packages
library(tidyverse)
library(reshape2)
library(treemapify)
library(waffle)

# Data
treemap_df <- tibble(group_name = c("True Pos.",
                                    "False Pos.",
                                    "True Neg.",
                                    "False Neg."),
                     frequency = c(47.5, 47.5, 902.5, 2.5),
                     test_result = c("Positive", 
                                     "Positive",
                                     "Negative", 
                                     "Negative"),
                     disease_status = c("ILL",
                                        "HEALTHY",
                                        "HEALTHY",
                                        "ILL"),
                     decision = c("Correct", 
                                  "Incorrect",
                                  "Correct",
                                  "Incorrect"))

pictogram_df <- tibble(group_name = c("True Pos.",
                                             "False Pos.",
                                             "True Neg.",
                                             "False Neg."),
                              frequency = c(9, 9, 81, 1),
                              test_result = c("Positive", 
                                              "Positive",
                                              "Negative", 
                                              "Negative"),
                              disease_status = c("Disease",
                                                 "Healthy",
                                                 "Healthy",
                                                 "Disease"),
                              decision = c("Correct", 
                                           "Incorrect",
                                           "Correct",
                                           "Incorrect"))
```

# Probability 101

## Probability vs. Statistics

Probability: Use known state of the world to predict the data we should observe.

* I have a fair coin (known state of the world) and can expect 50 heads and 50 tails, on average, from 100 flips (predicted data). 

Statistics: Use data we have observed to infer the most likely state of the world. 

* I flipped a coin 100 times and got 52 heads and 48 tails (observed data), I'm pretty sure that I have a fair coin (inferred state of the world). 

## Types of probability

Classic, based on symmetry. 

Frequentist, based on experience. 

Subjective, based on best guess. 

## Types of events

**Independent probabilities.** Should multiply. 

**Conditional probabilities.** Should use Bayes law. Visualise Bayesian updating with a contingency table/ infographic. 

## Puzzles

Sometimes probability is unintuitive. Birthday puzzle. 

# Legal problems

## Ignoring dependence

Sally Clark had two children dies from SIDS. Dr. Meadow, pediatrician, said there's 1 in 8543 chance of having a child die from SIDS giving a 1 in 73mil chance of having two children die from it. Potential genetic, environmental dependence ignored and she was sentenced for 3 years.

## The prosecutor's fallacy

**Prosecutor's fallacy:** False belief that conditional probabilities are invertible.

$$Pr(A|B) \neq Pr(B|A)$$

Forensics may compute that probability of blood at scene matching the defendent's by chance is 1%. That doesn't mean there's a 1% chance the defendent is innocent. We have been told the probability of a match given innocence, we want the probability of innocence given a match, but they don't equate:

$$Pr(\text{match}|\text{innocent}) \neq Pr(\text{innocent}|\text{match})$$

# Health problems

## Base rate neglect

Imagine there's a disease with 10% prevalence in the population. The screening test to identify the disease has 90% accuracy. You go to the doctor's and get a positive result on the test, what is the probability you have the disease?

Let's unpack the two parts of the scenario using an example population of 100 people.

Part 1: "The disease has a prevalence of 10%":

* 10% of the 100 people have the disease $\rightarrow$ 10 people
* 90% of the 100 people don't $\rightarrow$ 90 people

Part 2: "The test is 90% accurate":

* For the 10 people with the disease
    + 90% of them will be correctly identified as positive $\rightarrow$ 9 people
        - True positives
    + 10% of them will be incorrectly identified as negative $\rightarrow$ 1 person
        - False negatives
* For the 90 people without the disease
    + 90% of them will be correctly identified as negative $\rightarrow$ 81 people
        - True negatives
    + 10% of them will be incorrectly identified as positive $\rightarrow$ 9 people
        - False positives

Now we can use this breakdown to figure out the probability we're looking for. The probability you have the disease given you receive a positive result is just the fraction of all positive results that are true positives:

$$\begin{aligned}
\text{Pr}(D|+) &= \frac{\text{True pos.}}{\text{Total pos.}} \\
&= \frac{\text{True pos.}}{\text{True pos.} + \text{False pos.}} \\
&= \frac{9}{9 + 9} \\
&= 0.5
\end{aligned}$$

Visualising our example population makes it clear why the answer is 0.5. There are exactly the same number of positive results from the healthy portion of the population as there are from the diseased portion: 

```{r}
ggplot(pictogram_df, aes(label = disease_status, values = frequency, colour = test_result)) +
  geom_pictogram(n_rows = 9, flip = T) +
  scale_colour_manual(name = "Test Result",
                      values = c("Positive" = "Red", "Negative" = "Grey75")) +
  scale_label_pictogram(name = NULL,
                        values = c("Disease" = "user-md", "Healthy" = "smile")) +
  coord_equal() +
  theme(panel.background = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.key = element_blank())
```



# Research problems

# Statistical cognition

## The linda problem

Extra example of how we aren't intuitive statisticians. 

## All averages aren't made equal

Relative percentage change vs. absolute percentage change (i.e. percentage points) vs. natural frequencies. 