---
title: "Lizards playing rock, paper, scissors*"
date: "2019-08-28"
description: "An introduction to agent-based modelling and evolutionary simulation."
tags: ["Agent-based modelling", "Evolution"]
slug: abm_intro
output:
  blogdown::html_page:
    toc: true
    toc_depth: 1
    number_sections: true
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

```{r setup, include=FALSE}
# RMarkdown
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(igraph)
library(reshape2)
library(ggplot2)
```

This article was inspired by this [post](https://marcosmolla.com/2015/07/16/an-introduction-to-agent-based-modelling-in-r/), which I attempted to recreate and extend.  

If you want to see the full code for this post then you can find the .Rmd file [here](https://github.com/mitchelfruin/MitchThinks/blob/master/content/interests/abm_intro.Rmd).

# Rock, paper, scissors

Aim: Simulate simple rock, paper, scissors game to visualise a simple evolutionary process.  

Simple Game Rules:

* Individuals are organised on a lattice network.
* Each individual has a strategy (Rock, Paper, Scissors). 
* Each time period we choose one individual at random to play a game. 
* They can only play with one of their direct neighbors. 
* The loser takes on the strategy of the winner.

```{r simple function}
# Define function to simulate simple game 
simple_RPS <- function(n_rounds){
  
  # Define size of the lattice
  side_length <- 10 
  
  # Create an empty data.frame to store strategy frequencies
  strat_freq <- data.frame()
  
  # Create a lattice network using the igraph package
  l <- graph.lattice(length = side_length, dim = 2)
  
  # Every individual (vertices) chooses a strategy at random
  V(l)$weapon <- sample(c(1:3), size = length(V(l)), replace = T)
  
  # Colour the different options
  V(l)[weapon==1]$color <- "purple" # Paper
  V(l)[weapon==2]$color <- "blue" # Scissors (sapphire)
  V(l)[weapon==3]$color <- "red" # Rock
  
  # Initial arrangement: 
  plot(l,
       layout = as.matrix(expand.grid(1:side_length, 1:side_length)),
       vertex.label=NA)
  
  # Simulate n rounds
  for(t in 1:n_rounds){
    
    # Pick a random agent
    from <- as.numeric(sample(V(l), 1))
    # Find its neighbours
    nei <- neighbors(l, v = from, mode = 'all')
    
    # If there is only one weapon type left, then stop simulation 
    if(length(unique(V(l)$weapon))==1) {
      stop(paste(unique(V(l)$weapon),
                 'has won the game after',
                 t,
                 'rounds!'))
      } else {
        # If there are multiple weapon types then choose 1 of agent's neighbours
        to <- sample(nei, 1)
        from_to <- c(from, to)
        
        # Find both their weapon types
        w <- as.numeric(V(l)$weapon[from_to])
        
        # If both weapons are equal, nothing happens:
        if(w[1]==w[2]) {} else{
          
          # If one chooses Rock and the other Paper, Paper wins:
          if(max(w) == 3 && min(w) == 1) {
            V(l)$weapon[from_to[w==3]] <- "1" 
            }
          else{
            # For the two other cases, the better weapon wins:
            V(l)$weapon[from_to[w == min(w)]] <- V(l)$weapon[from_to[w == max(w)]]
            }
          }
        }
    
    # Record the frequency of each strategy for each round
    strat_freq <- rbind(strat_freq,
                        c(sum(V(l)$"weapon" == "1"),
                          sum(V(l)$"weapon" == "2"),
                          sum(V(l)$"weapon" == "3")))
    
    # Plot the lattice network every 25th round
    if(t %% 25 == 0){
      V(l)[weapon==1]$color <- "purple" # Paper
      V(l)[weapon==2]$color <- "blue" # Scissors
      V(l)[weapon==3]$color <- "red" # Rock
      
      plot(l,
           layout = as.matrix(expand.grid(1:side_length, 1:side_length)),
           vertex.label=NA)
      }
    }

# Viz changes over time 
names(strat_freq) <- c("Paper", "Scissors", "Rock")
s <- melt(strat_freq)
s$time <- 1:nrow(strat_freq)

ggplot(s, aes(x = time, y = value, col = variable)) + 
  geom_line() + 
  scale_y_continuous("Frequency") +
  scale_x_continuous("Time") +
  scale_color_discrete("Strategy") +
  theme_minimal() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
}
```

```{r simulate RPS, cache = TRUE}
# Play for 1000 rounds
simple_RPS(n_rounds = 1000)
```

# Side Blotched Lizards

Aim: Simulate a stylized version of common side-blotched lizard evolution to visualise a more complicated evolutionary process. 

SBL rules: 

* Individuals are organised on a large lattice network (some blank space).
* Each individual has a colour.  
* Each node has a food value.  
* They can only play with their direct neighbors. 
* Orange beats Blue, Blue beats Yellow, Yellow beats Orange. 
* If they lose, they take over the strategy of the winner. 

```{r SBL function, eval=FALSE}
# SBL function (unfinished)
SBL_sim <- function(n_rounds){
  
  # Define size of the lattice
  side_length <- 15 
  
  # Create an empty data.frame to store strategy frequencies
  strat_freq <- data.frame()
  
  # Create a lattice network using the igraph package
  l <- graph.lattice(length = side_length, dim = 2)
  
  # Every individual (vertices) chooses a strategy at random
  V(l)$weapon <- sample(c(1:4), size = length(V(l)), replace = T)
  
  # Colour the different options
  V(l)[weapon == 0]$color <- "white" # Vacant
  V(l)[weapon == 1]$color <- "yellow" # Sneakers
  V(l)[weapon == 2]$color <- "blue" # Dominant
  V(l)[weapon == 3]$color <- "orange" # Ultra Dominant
  
  # Initial arrangement: 
  plot(l,
       layout = as.matrix(expand.grid(1:side_length, 1:side_length)),
       vertex.label=NA)
  
  # Simulate n rounds
  for(t in 1:n_rounds){
    
    # Pick a random agent
    from <- as.numeric(sample(V(l), 1))
    # Find its neighbours
    nei <- neighbors(l, v = from, mode = 'all')
    
    # If there is only one weapon type left, then stop simulation 
    if(length(unique(V(l)$weapon)) == 1) {
      stop(paste(unique(V(l)$weapon),
                 'has won the game after',
                 t,
                 'rounds!'))
    } else {
      # If there are multiple weapon types then choose 1 of agent's neighbours
      to <- sample(nei, 1)
      from_to <- c(from, to)
      
      # Find both their weapon types
      w <- as.numeric(V(l)$weapon[from_to])
      
      # Find neighbours weapon types
      neigh_w <- as.numeric(V(l)$weapon[nei])
      
      # If chosen isn't blank and neighbour is same type, then take over any blank neighbours:
      if(w[1] != 0 & w[1] == w[2]) {
        V(l)$weapon[nei[w == 0]] <- w[1]
      } 
      else{
        # If they're different types, then competitive rules apply
        if(max(w) == 3 && min(w) == 1) {
          V(l)$weapon[from_to[w==3]] <- "1" 
        }
        else{
          V(l)$weapon[from_to[w == min(w)]] <- V(l)$weapon[from_to[w == max(w)]]
        }
      }
    }
    
    # Record the frequency of each strategy for each round
    strat_freq <- rbind(strat_freq,
                        c(sum(V(l)$"weapon" == "1"),
                          sum(V(l)$"weapon" == "2"),
                          sum(V(l)$"weapon" == "3"),
                          sum(V(l)$"weapon" == "0")))
    
    # Plot the lattice network every 25th round
    if(t %% 25 == 0){
      V(l)[weapon == 1]$color <- "yellow" # Sneakers
      V(l)[weapon == 2]$color <- "blue" # Dominant
      V(l)[weapon == 3]$color <- "orange" # Ultra Dominant
      V(l)[weapon == 0]$color <- "white" # Vacant
      
      plot(l,
           layout = as.matrix(expand.grid(1:side_length, 1:side_length)),
           vertex.label=NA)
    }
  }
  
  # Viz changes over time 
  names(strat_freq) <- c("Sneakers", "Dominant", "Ultra-Dominant", "Vacant")
  s <- melt(strat_freq)
  s$time <- 1:nrow(strat_freq)
  
  ggplot(s, aes(x = time, y = value, col = variable)) + 
    geom_line() + 
    scale_y_continuous("Frequency") +
    scale_x_continuous("Time") +
    scale_color_discrete("Type") +
    theme_minimal() +
    theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
}
```

```{r SBL sim, eval = FALSE}
SBL_sim(n_rounds = 500)
```

# Agent-based modeling

Summary. 

Use/relevance in complexity economics.