---
title: "Nobody goes there anymore. It's too crowded."
date: "2019-08-28"
description: "Introduction to the 'El-Farol Bar Problem', a staple of complexity economics."
tags: ["Agent-based modelling", "Complexity", "Economics"]
slug: el_farol_bar_problem
draft: true
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
    number_sections: true
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

```{r setup, include=FALSE}
# RMarkdown
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(tidyverse)
library(truncnorm)
```

I'm currently making my way through *Complexity and the Economy* by W. Brian Arthur. The book's second chapter is a paper from 1994 called [*Inductive Reasoning and Bounded Rationality: The El Farol Problem*](http://tuvalu.santafe.edu/~wbarthur/Papers/El_Farol.pdf). It's a fantastic example of how marrying psychological insights and a computational (rather than analytic) approach can produce novel, interesting, and intuive research in economics. So, I thought I'd have a go at using R to reproduce the model described in the paper.   

# Background 

When Arthur was working at the Santa Fe Institute in 1993, there was a bar called El Farol where there was Irish music on thursday nights. He noted that people would go if they expected few people to be there, but avoided it if they expected it to be too busy. This problem is interesting because it creates something of a paradox where beliefs lead to outcomes that cause those beliefs to be proved false. If people expect many others to attend, then few will show up and the bar will be empty. If people expect nobody to attend, then everyone will show up and it'll be packed. In both cases the original belief causes behaviour that means the belief ends up being incorrect.

Let's generalise this phenomenon in Arthur's toy model and see if we can gain any insight into the dynamics of attendance over time.  

# Model 

## Set-up

100 people decide independently each week whether or not to attend a bar that offers music on a certain night. The music is enjoyable if the bar isn't too busy, specifically if fewer than 60% of the possible 100 people attend. There's no way to tell in advance how many people will show up. So, each person goes if they expect fewer than 60 people to attend but stays home if they think more than 60 people will be there.

There's no collusion or prior communication between possible attendees and the choice isn't affected by whether the person was right or wrong in their beliefs about previous visits. The only info available is a list of the number of people who came in the past weeks.

For example, the past 10 weeks' attendance might be:

```{r attendance eg}
# Example attendance
set.seed(23)
round(rtruncnorm(10, 0, 100, 50, 20), 0)
```

Each of the 100 people can individually form $k$ hypotheses they could use to predict the next week's attendance based on the figures for the past $d$ weeks. For example, their hypotheses could include:

* Predict the same as last week $\rightarrow$ 82 people
* Predict a mirror image around 50 of last week $\rightarrow$ 28 people
* Predict the average of the past four weeks $\rightarrow$ 62 people

Each person decides to go out or stay in according to the currently most accurate prediction hypothesis in their individual set. 

## R function 

# Simulations

## If we start low

## If we start high

## Results

Mention ecology?

# Why should you care?

It exemplifies a common type of problem. 

It shows the value of numerical solutions to problems when analytical ones become too hard or are impossible (in this case because of inductive reasoning). 

Conclusion should relate to the Yogi Berra quote about his favourite restaurant in St. Louis, Ruggeri's. 
