---
title: "Estimating Pi 2000 years too late."
author: "Mitchel Fruin"
date: "2020-05-08"
description: "Comparing Archimedes' estimate of Pi with Rchimedes estimate."
tags: 
slug: estimate_pi
draft: true
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
    number_sections: true
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

```{r setup, include=FALSE}
# RMarkdown
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      fig.width = 8.09,
                      fig.height = 5)

# Packages
library(tidyverse)
library(scales)
library(plotly)
```

A couple of month's ago I read Steven Strogatz's Inifinite Powers.

It's a good book overall but the first half is great. I'd recommend it to anyone who ever asked, "but, like, why?" in a maths class in secondary school and didn't get a good answer (typically either a 'don't interrupt me' look or worse, "because it is"). 

My two favourite sections invovled Storgatz recounting how Archimedes discovered the formula for the area of a circle and how he estimated $\pi$. This post will only cover the latter. The reason being that there's a popular data science interview question which requires the applicant to estimate $\pi$. So, I thought it would be fun to compare Archimedes' estimate to what I'll be (shamelessly and repeatedly) calling **R**chimedes' estimate.

# Archimedes' estimate

# **R**chimedes' estimate

Ok, you're dressed to impress, you've politely said yes to a tea because you read a social psychology paper once that said interviewers preferred candidates that did, and you can feel a few beads of nervous sweat on the small of your back. You sit down at the interview table and the interviewer hands you a laptop with a set of instructions already open:

> You have a function which takes the input $n$ and outputs $n$ random numbers between 0 and 1. Estimate the value of $\pi$.

After the customary moment of panic, you decide actually think it through. 

Well, the easiest way you can think of to estimate $\pi$ is to find the area of a circle with a radius of $1$. 

After all,

$$\begin{aligned}
\text{Area of a circle} &= \pi \times \text{radius}^2 \\
\end{aligned}$$

So if $\text{radius} = 1$, then 

$$\begin{aligned}
\text{Area of a circle} &= \pi \times 1^2 \\
&= \pi
\end{aligned}$$

Alright, so what would a circle with radius of 1 look like?

```{r, fig.asp=1}
circleFun <- function(center = c(0, 0), diameter = 2, npoints = 100){
    r = diameter / 2
    tt <- seq(0, 2*pi, length.out = npoints)
    xx <- center[1] + r * cos(tt)
    yy <- center[2] + r * sin(tt)
    return(data.frame(x = xx, y = yy))
}

r_1 <- circleFun(npoints = 10000)

r_1 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 0.5, col = "purple") +
  geom_vline(aes(xintercept = 0)) +
  geom_hline(aes(yintercept = 0)) +
  theme_minimal()
```

```{r, fig.asp = 1}
random_xy <- tibble(x = runif(1000), 
                    y = runif(1000))

r_1 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 0.5, col = "purple") +
  geom_point(data = random_xy, aes(x = x, y = y)) +
  geom_vline(aes(xintercept = 0)) +
  geom_hline(aes(yintercept = 0)) +
  theme_minimal()
```

Find the area of a circle with radius 1.

```{r}
R_chimedes <- function(n_points) {
  # Generate random pair of coordinates between 0 and 1
  x <- runif(n_points)
  y <- runif(n_points)
  # For each pair find distance from the origin
  o_dist <- sqrt(x^2 + y^2)
  # If o_dist <= 1, then inside or on circle
  in_or_out <- o_dist <= 1
  # Estimate pi
  4 * mean(in_or_out == TRUE)
}
```

How many random points do we need before we are more accurate than Archimedes?

Lets try a few different powers of 10.

```{r}
set.seed(1)
estimates <- tibble(n_points = 10^(2:7),
                    attempt_1 = map_dbl(n_points, R_chimedes),
                    attempt_2 = map_dbl(n_points, R_chimedes),
                    attempt_3 = map_dbl(n_points, R_chimedes))

estimates %>%
  pivot_longer(cols = c(attempt_1, attempt_2, attempt_3),
               names_to = "Estimate") %>%
  ggplot(aes(x = n_points, y = value, col = Estimate)) +
  geom_point(aes(shape = Estimate)) +
  geom_hline(aes(yintercept = pi)) +
  geom_line(aes(lty = Estimate)) +
  scale_x_log10(breaks = 10^(2:7),
                labels = trans_format("log10", math_format(10^.x))) +
  scale_color_viridis_d() +
  theme_minimal()

estimates %>%
  pivot_longer(cols = c(attempt_1, attempt_2, attempt_3),
               names_to = "Estimate") %>%
  ggplot(aes(x = n_points, y = value, col = Estimate)) +
  geom_point(aes(shape = Estimate)) +
  geom_hline(aes(yintercept = pi)) +
  geom_line(aes(lty = Estimate)) +
  scale_x_log10(breaks = 10^(5:7),
                labels = trans_format("log10", math_format(10^.x)),
                limits = c(10^5, 10^7)) +
  scale_y_continuous(limits = c(3.135, 3.145)) +
  scale_color_viridis_d() +
  theme_minimal()
```

```{r}
estimates %>%
  mutate(Pi = pi) %>%
  pivot_longer(cols = c(attempt_1, attempt_2, attempt_3, Pi),
               names_to = "Estimate") %>%
  plot_ly(x = ~n_points,
          y = ~value,
          color = ~Estimate,
          colors = viridis_pal(direction = -1)(4),
          text = ~paste0(Estimate, ": ", value),
          hoverinfo = "text") %>%
  add_markers(symbol = ~Estimate,
              showlegend = FALSE) %>%
  add_lines(showlegend = FALSE) %>%
  layout(xaxis = list(title = "Number of Points", type = "log"),
         yaxis = list(title = "Value"))
```

# Knowing vs. Understanding

About undersatdning vs. knowing.

For a long time, definitely throughout all of school, all I really did was *know* stuff. Now, I want to *understand* stuff. And even if that means there's less stuff involved, I'm fine with that. Value of understanding is that you begin to see connections between seemingly disparate points of *knowledge*.