---
title: "Probability in the real world*"
author: "Mitchel Fruin"
date: "2019-08-15"
description: "Article with real world examples of probability mistakes."
tags: ["Probability & Statistics"]
slug: probability_real_world
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
    number_sections: true
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

```{r setup, include=FALSE}
# Rmarkdown
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# Packages
library(tidyverse)
library(reshape2)
library(treemapify)
library(waffle)

# Data
treemap_df <- tibble(group_name = c("True Pos.",
                                    "False Pos.",
                                    "True Neg.",
                                    "False Neg."),
                     frequency = c(47.5, 47.5, 902.5, 2.5),
                     test_result = c("Positive", 
                                     "Positive",
                                     "Negative", 
                                     "Negative"),
                     disease_status = c("ILL",
                                        "HEALTHY",
                                        "HEALTHY",
                                        "ILL"),
                     decision = c("Correct", 
                                  "Incorrect",
                                  "Correct",
                                  "Incorrect"))

pictogram_df <- tibble(group_name = c("True Pos.",
                                             "False Pos.",
                                             "True Neg.",
                                             "False Neg."),
                              frequency = c(9, 9, 81, 1),
                              test_result = c("Positive", 
                                              "Positive",
                                              "Negative", 
                                              "Negative"),
                              disease_status = c("Disease",
                                                 "Healthy",
                                                 "Healthy",
                                                 "Disease"),
                              decision = c("Correct", 
                                           "Incorrect",
                                           "Correct",
                                           "Incorrect"))
```

# Linda's birthday

## The Linda problem

## The Birthday puzzle

## Bad intuitions

These are just two examples from many which illustrate how our gut reactions can lead us astray. 

But these examples were artificial, and contrived, and designed to flummox us, so why should you care about the odd mistaken probability? Well, it could save your life.

# Legal problems 

Uncertainty is inherent in the process of a courtroom trial. This means that crucial issues in a trial can hinge on questions of probability, for example:

* How plausible is the version of events presented by the defence? 
* What is the probability that the blood at the scene would match the defendent by chance? 
* How should the jury incorporate new evidence into their pre-existing beliefs about a defendent's guilt or innocence? 

Whether through genuine ignorance or manipulation, incorrect statements about probability when the stakes are so high can have dire consequences. 

## Ignoring dependence

In 1996, the infant son of Sally Clark, a solicitor in Manchester, died suddenly when he was 11 weeks old. She had another baby boy the following year who tragically also died, this time at 8 weeks old. Suspecting the two deaths weren't a coincidence the police charged Clark with two counts of murder. As an expert witness for prosecution, the pediatrician Dr. Roy Meadow claimed it was vanishingly unlikely that two children in the same family would die from Sudden Infant Death Syndrome (SIDS). He estimated a 1 in 8543 chance of a representative child dying from SIDS and testified that this gave a 1 in 73 million chance of having two children die from it. Clark was sentenced to life in prison.

There are a variety of problems with his calculation but we'll focus on the easiest one first: Dr. Meadow assumed the two events were independent. 

Potential genetic, environmental dependence ignored and she was sentenced for 3 years.

Discuss independence vs. dependence here. 

Introduce Bayes' rule. 

## The prosecutor's fallacy

**Prosecutor's fallacy:** False belief that conditional probabilities are invertible.

$$Pr(A|B) \neq Pr(B|A)$$

Forensics may compute that probability of blood at scene matching the defendent's by chance is 1%. That doesn't mean there's a 1% chance the defendent is innocent. We have been told the probability of a match given innocence, we want the probability of innocence given a match, but they don't equate:

$$Pr(\text{match}|\text{innocent}) \neq Pr(\text{innocent}|\text{match})$$

## The Feynman conjecture 

Paintball analogy. 

Losing the lottery: Dutch nurse example. 

# Health problems

Problems in medicine. 

## Base rate neglect

Imagine there's a disease with 10% prevalence in the population. The screening test to identify the disease has 90% accuracy. You go to the doctor's and get a positive result on the test, what is the probability you have the disease?

Let's unpack the two parts of the scenario using an example population of 100 people.

Part 1: "The disease has a prevalence of 10%":

* 10% of the 100 people have the disease $\rightarrow$ 10 people
* 90% of the 100 people don't $\rightarrow$ 90 people

Part 2: "The test is 90% accurate":

* For the 10 people with the disease
    + 90% of them will be correctly identified as positive $\rightarrow$ 9 people
        - True positives
    + 10% of them will be incorrectly identified as negative $\rightarrow$ 1 person
        - False negatives
* For the 90 people without the disease
    + 90% of them will be correctly identified as negative $\rightarrow$ 81 people
        - True negatives
    + 10% of them will be incorrectly identified as positive $\rightarrow$ 9 people
        - False positives

Now we can use this breakdown to figure out the probability we're looking for. The probability you have the disease given you receive a positive result is just the fraction of all positive results that are true positives:

$$\begin{aligned}
\text{Pr}(D|+) &= \frac{\text{True pos.}}{\text{Total pos.}} \\
&= \frac{\text{True pos.}}{\text{True pos.} + \text{False pos.}} \\
&= \frac{9}{9 + 9} \\
&= 0.5
\end{aligned}$$

Visualising our example population makes it clear why the answer is 0.5. There are exactly the same number of positive results from the healthy portion of the population as there are from the diseased portion: 

```{r}
ggplot(pictogram_df, aes(label = disease_status,
                         values = frequency,
                         colour = test_result)) +
  geom_pictogram(n_rows = 9, flip = T) +
  scale_colour_manual(name = "Test Result",
                      values = c("Positive" = "Red", "Negative" = "Grey75")) +
  scale_label_pictogram(name = NULL,
                        values = c("Disease" = "user-md", "Healthy" = "smile")) +
  coord_equal() +
  theme(panel.background = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.key = element_blank())
```

This is because the 10% inaccuracy that generates false positive results is applied to the larger segment of the population (the 90 people who are healthy). The tendency to overestimate this probability is referred to as 'base rate neglect' because we're under-weighting the base probability of having the disease, 10%, when calculating our answer. 

We can also do it the 'proper' way using Bayes' rule discussed previously:

add Bayes here

This isn't a contrived or unimportant example. Breast cancer example. 

# Research problems

Problems in research and how research is communicated. 

## p-value problems 

## Changes in probability

Relative percentage change vs. absolute percentage change (i.e. percentage points) vs. natural frequencies. 